---
title: "Climate Change Module"
author: 'Charlie Keglovitz'
format:
  html:
    embed-resources: true
---

```{r message=FALSE}
library(tidyverse)
```

## Warm-up: Examining CO2 trends in R

- Example from <http://climate.nasa.gov/vital-signs/carbon-dioxide/>
- Raw data from <https://gml.noaa.gov/webdata/ccgg/trends/co2/co2_mm_mlo.txt>

In 1958, Dr. Charles David Keeling (1928-2005), a scientist at Scripps Institute of Oceanography, began collecting data on atmospheric CO2 concentration at the Mauna Loa Observatory located in Hawaii. This dataset allowed us to understand the degree to which climate change is human-caused through our burning of fossil fuels and release of CO2 into the atmosphere. Due to his scientific achievements, Dr. Keeling was awarded the National Medal of Science by President George W. Bush in 2002. This is the highest award for lifetime scientific achievement that can be granted in the U.S. Today, you get to analyze this same dataset, except that you have more data than was available to Dr. Keeling and his colleagues because your dataset extends up to the current time.

To read the code, you will use a new function called `read_table`.  It is similar to `read_csv` except it looks for spaces between column entries rather than commas (remember that csv stands for comma-separated values).  Others include `read_tsv`, which uses tabs to separate entries.  You can discover the separation type by putting <https://gml.noaa.gov/webdata/ccgg/trends/co2/co2_mm_mlo.txt> into your web browser and examining the file.  The code also includes an argument for comment, which denotes the character used to define a line in the file as being a comment rather than data or a header.  The file also reveals that the column headers are on lines with a comment character, so they won't be read.  You will use the argument `col_names` to define the headers manually.  Finally, `NA` values are defined by the numbers -1 and -99, so they need to be defined using the `na` argument (otherwise, they will be read as numeric values).

```{r message=FALSE}

co2 <-  read_table("https://gml.noaa.gov/webdata/ccgg/trends/co2/co2_mm_mlo.txt", 
                  comment="#",
                  col_names = c("year", "month", "decimal_date",
                                "monthly_average",
                                "deseasonalized", "days", "sd_days",
                                "unc_month_mean"),
                  na = c("-1", "-99.99"))
co2
```


```{r}
ggplot(co2, aes(x = decimal_date, y = monthly_average)) + 
  geom_line() + 
  geom_line(aes(y = deseasonalized), color = "blue") +
  labs(x = "Year", y = "CO2 concentration (ppm)")
```

**Question 1:**

Describe the overall trend in the CO2 data.

**Answer 1:**

The general trend over the entire time period is a steady increase in CO2 concentration. The slope of the increase seems to have increased slightly since 2000 but it seems relatively linear in growth.

**Question 2:**

How does CO2 vary within a year?  What month is it at max?  Min?  What might explain this sub-annual pattern? (you will need to write code and make a plot to answer this question)

**Answer 2:**

The CO2 data oscillates every year from high in the winter to lower in the summer. It is at its max in May and very close to the peak in June. CO2 is lowest during October. I think trees photosynthesizing carbon dioxide into oxygen more during the spring/summer would lower CO2 levels and they would subsequently rise as tress die.

## Global Temperature Data

Current climate change affects many aspects of the environment, with socio-economic consequences. For example, a warmer climate can allow new diseases to be introduced and persist (e.g., West Nile became established in the United States after an unusually warm winter, which allowed the mosquitoes that carry the virus to survive and spread). We are concerned not only with the actual temperature but also with the rate at which it changes. Rapid changes make it more likely that species cannot adapt and will go extinct. 

Each of the most recent years has been the warmest on record. In this section, we will analyze global mean temperature data.

Data from: <https://climate.nasa.gov/vital-signs/global-temperature/>

**Question 3:**

Describe the data set to the best of your ability, given the documentation provided. 

- Where is the data from?
- Describe the class of each column and what units it is measured in. 
- What is the difference between "global temperature" and "global temperature anomaly"? 

**Answer 3:**

1) The data is the Global Land-Ocean Temperature Index from NASA, specifically the Goddard Institute for Space Studies.

2) The class of each column is a double. year, month, and days are measured in years, months, and days. Decimal date is the amount of years that compiles the year and how far along it is in the year using months. Monthly average and deseasonalized have the units of co2 parts per million. sd_days is a measure of standard deviation and unc_month_mean is a measure of mean.

3) Global temperature is the standard temperature while global temperature anomaly is how much warmer or colder a year/month was compared to the long running average or the temperature that it was expected to be.

**Question 4:**

Construct the necessary R code to import and prepare for plotting the following data set: <https://data.giss.nasa.gov/gistemp/graphs/graph_data/Global_Mean_Estimates_based_on_Land_and_Ocean_Data/graph.txt>

You'll need to determine the file's delimiter and any comments or skips. You will also need to be sure that you have column names. You will not directly use the code above to read the CO2 data, but that code provides helpful tips for reading the data.

**Answer 4:**

```{r}

tempIndex <- read_table("https://data.giss.nasa.gov/gistemp/graphs/graph_data/Global_Mean_Estimates_based_on_Land_and_Ocean_Data/graph.txt", 
                  skip = 5,
                  col_names = c("Year", "No_Smoothing", "Lowess(5)"))

```


**Question 5:**

Plot the trend in global mean temperatures over time.  Describe what you see in the plot and how you interpret the patterns you observe.

**Answer 5:**

```{r}

tempIndex |> 
  ggplot(aes(x = Year, y = No_Smoothing)) +
  geom_line() + 
  geom_point()

```
In the plot I see that temperature levels fluctuated but did not rise greatly overall from 1880 to just before 1980. Afterwards there was still fluctuation year to year but an overall steady rise in temperature that has only increased more drastically as time went on. I interpret this result as temperature staying relatively stable as the world emmited fewer greenhouse gases but then steadily rising as we emmited more of these gases.


## Evaluating the evidence for a "Pause" in warming?

The [2013 IPCC Report](https://www.ipcc.ch/pdf/assessment-report/ar5/wg1/WG1AR5_SummaryVolume_FINAL.pdf) included a tentative observation of a "much smaller increasing trend" in global mean temperatures since 1998 than was observed previously.  This led to much discussion in the media about the existence of a "Pause" or "Hiatus" in global warming rates, as well as much research looking into where the extra heat could have gone.  (Examples discussing this question include articles in [The Guardian](http://www.theguardian.com/environment/2015/jun/04/global-warming-hasnt-paused-study-finds), [BBC News](http://www.bbc.com/news/science-environment-28870988), and [Wikipedia](https://en.wikipedia.org/wiki/Global_warming_hiatus)). 

You will use rolling averages to help you explore the evidence for a pause. Since you have not been provided instructions for calculating rolling means, the learning objective of this question is to practice finding the solution.

**Question 6:**

Use a search engine (e.g., Google) or a chat LLM to find out how to calculate a rolling average in R. What search term or chat LLM prompt did you use?  What website or answer from the chat LLM did you end up using?  How much did you need to modify the code from the website or chat LLM to answer the question successfully?

**Answer 6:**

```{r}

library(zoo)

tempIndex <- tempIndex |> 
  mutate(Three_year_rolling_avg = rollmean(No_Smoothing, k = 3, fill = NA, align = "center"))


```
I used ChatGPT to answer this question. I told it that I had a dataframe and wanted to calculate the rolling average of one of the columns and put that data into a new column. It told me to mutate using the rollmean function and to align each rolling mean at the center which means the first and last points will be NA. I did not have to modify the answer it gave me much for it to work.

**Question 7:**

- What is the meaning of "5-year average" vs. "annual average"?
- Create a data frame from the annual temperature anomaly data (from Question 4) with three new columns: 5-year running averages, 10-year running averages, and 20-year running averages.

**Answer 7:**
1) The meaning of a 5-year average is the average over only the last 5 years and not considering anything before that. An annual average only takes into account the last year.
2) 
```{r}

rollingAvg_tempIndex <- tempIndex |> 
  mutate(Five_year_rolling_avg = rollmean(No_Smoothing, k = 5, fill = NA, align = "center")) |> 
  mutate(Ten_year_rolling_avg = rollmean(No_Smoothing, k = 10, fill = NA, align = "center")) |> 
  mutate(Twenty_year_rolling_avg = rollmean(No_Smoothing, k = 20, fill = NA, align = "center"))

```

**Question 8:**

Plot the different averages on the *same plot* and describe what differences you see and why.  

**Answer 8:**

```{r}

rollingAvg_tempIndex |> 
  ggplot(aes(x = Year)) +
  geom_line(aes(y = No_Smoothing, color = "Single-Year")) +
  geom_line(aes(y = Five_year_rolling_avg, color = "5-Year Avg")) +
  geom_line(aes(y = Ten_year_rolling_avg, color = "10-Year Avg")) +
  geom_line(aes(y = Twenty_year_rolling_avg, color = "20-Year Avg"))


```

I see that the 20 year average is more resistant to sharp fluctuations year to year. The longer time intervals describe the overall trend while the shorter intervals are more vulnerable to fluctuations in the data on a yearly basis.

**Question 9:**

By examining the data here, what evidence do you find or not find for such a pause? 

**Answer 9:**

I do not find evidence of a pause since the five, ten, and twenty year averages continued to increase at a steady rate through 1998 and the early 200s. While the single year averages did have many spikes and drops that didn't seem to lead to a large overall increase, the trend of increasing temperatures since the late 1960s continued throughout that time period.

## Longer term trends in CO2 Records

When analyzing Earth’s climate, it is important to remember that Earth is 4.54 billion years old. Our analyses so far have only looked at recent history. How can we compare the recent data to prehistoric times? Are the current rates of change similar or different from those the earth has experienced in the past? To explore this, we can use data from ice cores drilled at the poles. 

Hundreds of ice cores have been extracted from polar ice because they contain valuable data on atmospheric chemistry over pre-historic times. These valuable data exist in tiny air bubbles trapped in the ice. These air bubbles contain the same gases in the same ratios as the atmosphere at the time when the ice formed. The data you will analyze today are from ice cores extracted from the Vostok research station in Antarctica. As you have probably assumed, the depth of the ice core is related to how old the ice is; deep ice is older. There are two other variables that you will be analyzing from the ice cores.  You will analyze CO2 concentration, which has been measured from air bubbles trapped in the ice. We can use these data to see what rates of change were like during this pre-historic period, during which human activity was minimal. 

[Ice core data](https://data.ess-dive.lbl.gov/view/doi%3A10.3334%2FCDIAC%2FATG.009):

Vostok Core, back to 400,000 years before the present day 

- Description of data set: <https://data.ess-dive.lbl.gov/view/doi%3A10.3334%2FCDIAC%2FATG.009>
- data: <https://data.ess-dive.lbl.gov/catalog/d1/mn/v2/object/ess-dive-457358fdc81d3a5-20180726T203952542>

You will use this code to download the data to your computer.

```{r}
download.file("https://data.ess-dive.lbl.gov/catalog/d1/mn/v2/object/ess-dive-457358fdc81d3a5-20180726T203952542",
              destfile = "vostok.icecore.co2")
```

You can click on the file in your Files pane to view it before reading into R.


**Question 10:**

The broad question is: how do recent atmospheric CO2 levels compare to historical levels?

Your answer to Question 10 is going to be a mix of code chunks and text that you put below in "Answer 10:"

- Describe the data set: what are the columns and units? Where do the numbers come from? 
- What is the temporal resolution of the data? 
- Read in and prepare data for analysis.   
- Reverse the ordering to create a chronological record so that each measurement is associcated with calender year.
- Plot data.
- Combine this time series with the Mauna Loa data (the Warm-up exercise in this assignment).  You will use "bind_rows()".
- Plot the combined data. (the most recent time period must be on the right side of the plot).        
- Describe your conclusions to the question "How do recent atmospheric CO2 levels compare to historical levels?" using your plot as supporting evidence.   

**Answer 10:**
1) The columns of this data set are depth of the ice, age of the ice, age of the air, and co2 in the ice. The units of depth is in meters, units of age of ice is in years, age of air is in years, and co2 concentration is in co2 parts per million. These numbers are coming from ice samples from the arctic to measure historical co2 levels trapped in the ice. These values do not change considerably over time so it is a good measure of historical co2 levels.

2) After doing some research, the temporal resolution of the ice is clearest at the top where the ice is newer but gets less clear as the ice gets deeper. Because of the method of measuring the data, we can not get precise year to year co2 measurements from prehistoric times, but we can get a general sense of how atmospheric co2 was changing over periods of hundreds to thousands of years.

3) 
```{r}
library(readr)

iceCore <- read_table2("vostok.icecore.co2", 
                        col_names = c("Depth_m", "Age_Ice_BP", "Age_Air_BP", "CO2_ppmv"),
                        col_types = "didd",
                        comment = "*")
iceCore <- na.omit(iceCore)
summary(iceCore)
str(iceCore)
```

4) 
```{r}

iceCore <- iceCore |> 
  arrange(desc(Age_Ice_BP))

```

5) 
```{r}

iceCore |> 
  ggplot(aes(x = Age_Ice_BP, y = CO2_ppmv)) + 
  geom_line() + 
  geom_point() +
  scale_x_reverse() +
  labs(x = "Age of Ice (Years)", y = "CO2 concentration (ppm)")

```
6)
```{r}
iceCore <- iceCore |> 
  mutate(decimal_date = 2025 - Age_Ice_BP) |> 
  mutate(monthly_average = CO2_ppmv)

combined_df <- bind_rows(co2, iceCore)

```

7) 
```{r}
combined_df |> 
  ggplot(aes(x = decimal_date, y = monthly_average)) + 
  geom_line() + 
  labs(x = "Year", y = "CO2 concentration (ppm)")

```
8) CO2 levels right now are higher than they have ever been in history. In prehistoric times, there were many peaks and valleys of co2 concentration, but they never reached as high as 300 ppm, let alone 400 ppm where levels are above now. This is a recent trend that does not follow the patterns of co2 concentration found in ice in the hundreds of thousands of years before human interference.

# Render and committing

Remember to Render your document as HTML and comment+push to GitHub your code and rendered HTML that was created when you knitted the document.  Your GitHub repository should have multiple commits with informative commit messages.

# Attribution

Include citation of any AI-generated assistance or discussion with classmates (per policy in syllabus). Proper documentation of AI-generated assistance includes the prompt, the source (e.g., ChatGPT), and the significant parts of the response.  Proper documentation of discussion with classmates include listing their names and the components discussed.  



